---
title: "GIS analyses with squirrygis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GIS analyses with squirrygis}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

# Background

The world of GIS software is vast and dark.
The software and utilities out there are powerful, but written in the language of Mordor.
Calling them often requires to learn new frameworks, find the lost documentation, and prototype endless code that will never be used.
Even if you succeed, your ears will blast and your soul will be lost.
As I found that most of the commonly used routines are quite easy to implement, I started implementing them in C++ for speed.
I then packaged them and the end result was squirrygis.
squirrygis is still a work in progress: I add new routines as I need them.
Currently, squirrygis is tailored to calculate bioclimatic variables for monthly timeseries data.
The available bioclimatic variables are:

  - BIO10

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(squirrygis)
library(terra)
```

# Average temperature of the warmest quarter (BIO10)

Let's write a function to generate some fake raster data for temperature.
This function should take the dimension of the output raster and the minimum and maximum range of temperature.

```{r temperature-fn}
temperature <- function(rows, cols, min = -20, max = 30) {
  ans <- matrix(runif(rows * cols, min, max), nrow = rows, ncol = cols)
  ans <- rast(ans)
  return (ans)
}
```

We can use this to generate a monthly timeseries data.

```{r temperature-series}
series <- c(
  temperature(100, 100, -20, -10),
  temperature(100, 100, -15, -5),
  temperature(100, 100, -10, 0),
  temperature(100, 100, -5, 5),
  temperature(100, 100, 0, 10),
  temperature(100, 100, 5, 15),
  temperature(100, 100, 10, 20),
  temperature(100, 100, 7, 17),
  temperature(100, 100, 4, 14),
  temperature(100, 100, 1, 11),
  temperature(100, 100, -2, 8),
  temperature(100, 100, -5, 5)
)
names(series) <- seq_len(12)
plot(series[[1]], col = heat.colors(100))
```

It is mandatory for the raster stack of timeseries to have 12 layers, one per month.
This is a strict requirement, but a necessary one.
In fact, most of the bioclimatic variables will be not accurate if only one layer is missing.
You can name the stack layers with names, as I did, or not.
If the names are different from 1...12 or are not in the correct order (e.g., May before July), squirrygis will return the warning:

```{r warn, echo=FALSE}
warning("Raster stack doesn't have valid names or they are not in the correct order")
```

In fact, this could have been an error instead of a warning.
I just assumed you know a bit what you're doing and be wise enough to pass the stack layers in the correct calendar order.
I would have been too paternalistic, otherwise, I think.

To calculate BIO10 is then straightforward:

```{r bio10}
b10 <- bio10(series)
plot(b10, col = heat.colors(100))
```

As quarters are defined as "three consecutive calendar months", I also implemented a way to check when this quarters starts:

```{r bio10-q}
q10 <- bio10(series, also.quarter = TRUE) #this returns a stack
b10 <- q10[[1]] #the first layer is BIO10
plot(q10[[2]]) #the second layer is the starting month of the quarter
```

I haven't seen this implemented before, but is quite insightful when doing analyses at large spatial scales.
You may want to *know* if the warmest quarter starts in different months in adjacent areas.
